Himalayan Expeditions Success Analysis

### 2024-04-18 paper/paper.qmd
Data section completed

The three datasets I considered included information about all peaks, all expeditions on those peaks and all members on those expeditions. The data from these three datasets are combined to form the main analysis dataset.

A person becomes an entry in my analysis dataset if, between 1905 and Spring 2019, they attempted to climb any one of the many Himalayan peaks in Nepal. It also included expeditions to both sides of border peaks as mentioned before.

@@

- `Height` is the height range in which the peak's heigh in metres falls. This is for the peak the person in the current entry is on an expedition for. The categoies for this are 5400 - 5749, 5750 - 6099, 6100 - 6449, 6450 - 6799, 6800 - 7149, 7150 - 7499, 7500 - 7849, 7850 - 8199, 8200 - 8549 and 8550 - 8900.
- `Seasons` is the season the expedition is in. This takes on either of the 4 values: Autumn, Spring, Winter, Summer.
- `Sex` is the sex reported by the expedition member and it is either male or female.
- `Age` is the age group in which the expedition member fell in at the time of the expedition. Depending on the best available data, this could be as of the summit date, the date of death, or the date of arrival at basecamp. The different categories for this are Under 18,19-30, 31-40, 41-50, 51-60, 61-70, 71-80 and 81-90.
- `Success`represents whether the person was successful in summitting the goal peak.
- `Solo` represents whether the person attempted a solo ascent.
- `Died` represents whether the person died during the expedition.

@@

@fig-counts shows

@@

@@

@fig-proportions shows

Out of these, I will be using `height of the peak`, `seasons`, `sex`, `age` and `solo` in the model as the independent variables and `success` as the dependent variable.

@@

For my analysis, a logistic regression model will be used to model if the person will be successful on their attempt to summit the particular Himalayan peak. The model will be based on five independent demographic variables: `height of the peak`, `sex`, `age`, `seasons` and `solo`.

>>>>

The three datasets I considered included information about all peaks, all expeditions on those peaks and all members on those expeditions. The data from these three datasets are combined to form the main analysis dataset. 
A person becomes an entry in my analysis dataset if, between 1905 and Spring 2019, they attempted to climb any one of the many Himalayan peaks in Nepal. It also included expeditions to both sides of border peaks as mentioned before.

@@

- `Height` is the height range in which the peak's height in metres falls. This is for the peak the person in the current entry is on an expedition for. The categories for this are 5400 - 5749, 5750 - 6099, 6100 - 6449, 6450 - 6799, 6800 - 7149, 7150 - 7499, 7500 - 7849, 7850 - 8199, 8200 - 8549 and 8550 - 8900.
- `Seasons` is the season the expedition is embarked in. This takes on either of the four values: Autumn, Spring, Winter, Summer.
- `Sex` is the sex reported by the expedition member and it is either male or female.
- `Age` is the age group in which the expedition member fell in at the time of the expedition. Depending on the best available data, this could be as of the summit date, the date of death, or the date of arrival at basecamp. The different categories for this are Under 18,19-30, 31-40, 41-50, 51-60, 61-70, 71-80 and 81-90.
- `Success`represents whether the person's expedition resulted in a successful summit.
- `Solo` represents whether the person attempted a solo ascent.
- `Died` represents whether the person died during the expedition.

@@

@fig-counts shows the counts for different variables we are considering to model. We see that there are significantly higher men than women. This shows the existence of a gender imbalance skewed towards men in mountaineering in the Himalayas. We also see other trends like most of the expedition members choose to not do solo ascents, most of the members fall in the middle-aged range and the peak expedition seasons are the more pleasant autumn and spring compared to the extreme seasons like winter and summer. These difference in counts need to be kept in mind when analysing success proportions.

@@

  plot_data$success <- factor(plot_data$success, levels = c(TRUE, FALSE))

@@

@fig-proportions shows the proportion of expedition members who has a successful outcome or died during their expedition. For sex, we see more men had successful expeditions compared to women. This is interesting keeping in the mind the vast difference in the counts of these two sexes going on expeditions. It might spark an interesting discussion into the gender imbalance in mountaineering. Additionally, we see most of the people who go on expeditions solo have more deaths which might explain the difference in the counts of people attempting solo ascent versus not attempting a solo ascent which was observed earlier. Some other trend we observe are that success rates decline with age and winter season contributes to the most deaths. The success and death proportions based on the height are very varied and do not follow a specific trend. This goes to show that just the height of the peak alone doesn't define the probability of having a successful summit but there are other factors which go into it.

@@

For my analysis, a logistic regression model will be used to model if the person will be successful on their attempt to summit the particular Himalayan peak. The model will be based on five independent demographic variables: `height of the peak`, `sex`, `age`, `seasons` and `solo` and the dependent variable will be `success`.
